On the Synchronization Power of the Openflow Data Plane

### 2015-03-29 inband-sync.tex
stefan sync

\algblockdefx[Bundle]{startTxn}{endTxn} %
[0]{\textbf{atomic}\{} %
[1][res]{ \} $\rightarrow #1$}

@@

\begin{abstract}
Control planes of modern Software-Defined Networks (SDNs)
are \emph{distributed systems}: to ensure availability and fault-tolerance,
to improve load-balancing, and to reduce overheads,
control modules are physically distributed.
%a redundant control plane
%can ensure a high availability, and a spatially distributed
%control plane allows to handle time-critical dataplane
%events close to their origins.
%[[PK this is a contraversial point, not important for us
%Moreover, an SDN may be managed by multiple administrators simultaneously.
%]]
However, a distributed control plane also introduces
new challenges, and in order to guarantee a consistent network operation,
the actions of different controllers may need to be synchronized.
%[[PK sounds contradicting to the fact that such systems already exist
%However, not much is known today about how to implement
%such a distributed control plane.
%]]
%
A well-known concept to solve synchronization and consistency problems
in distributed systems are \emph{transactions}: a transaction defines a sequence
of operations which are performed in an atomic, serializable manner, providing all-or-nothing semantics.
This paper shows that powerful transactions can also be implemented in software-defined networks,
using the standard Openflow protocol.
%Indeed, recent Openflow versions already come with useful but primitive
%features, in particular \emph{bundling} and \emph{rule-overlap checks}.
In particular, we present a
transactional interface which supports arbitrary atomic
transactions
over the switch configuration space: Our transactions
can include complex notions of conflicts (beyond flow overlap checks),
as well as computations and controller interactions. We discuss different use cases
for our interface (consistent policy installation, load-balancing, consensus),
and also describe how to
perform basic forms of policy compositions in-band.
\end{abstract}

@@

While the logically centralized perspective offered by SDN is attractive,
there is a wide consensus that
the control plane should be physically \emph{distributed}.
First, in order to provide a high availability as well as a minimal degree of
fault-tolerance, controllers should be redundant~\cite{onix,stn,onos}: a failure
of one controller can be masked by other controllers. Second, it has also been proposed
to distribute controllers \emph{spatially}, in order to handle latency-sensitive and
communication intensive control plane events close to their origin.~\cite{devoflow,kandoo,jukka,disco}
Third, larger SDNs are likely to be operated by multiple administrators or may even offer
participatory interfaces where different users can install and trigger policy changes
concurrently~\cite{participatory,stn}.
All this, however, comes with the need for \emph{synchronizing}
the distributed controllers:
the network should be manipulated \emph{consistently}~\cite{cpc}.
This task is nontrivial~\cite{cap-theorem}.

@@

Today, we do not have a good understanding yet of how to realize
such distributed control planes. The problem is essentially a
distributed systems
one: Multiple controllers may simultaneously try to
install conflicting updates and we want to resolve these conflicts
\emph{consistently} (no undesired behavior is observed on the data
plane) and \emph{efficiently} (no undesired delays are imposed on the
control application).

@@

This paper investigates the design of a transactional interface for software-defined
networks. In particular, we argue that the switch configuration state can be seen
as a shared memory which is modified concurrently by multiple controllers.
Being able to provide more complex and atomic sequences of transactions over this configuration
space can be appealing to resolve synchronization problems.

\paragraph{Our Contributions}
This paper introduces a very powerful transactional interface
 for software-defined networks. This interface is implemented
 in-band, and
 using Openflow~\cite{of-spec} (the standard SDN protocol today) features only.
Indeed, we find that Openflow's \emph{bundle} or \emph{rule overlap check}
features can serve as synchronization primitives.
However, without further mechanisms, their use is limited:
with a bundle alone it is not possible to react to
and change switch configuration state in any non-trivial way,
and with the rule overlap check it is not possible to react to
update conflicts defined over non-overlapping flow spaces.

In contrast, the transactional interface presented in this paper supports very general atomic
transactions over the switch configuration space, similar to classic transactional memory~\cite{stm-st95,tm-book}.
For instance, a transaction allows to modify switch configurations in an interactive
and computational fashion, as well as to define complex notions
of conflicts.
Our implementation is based on distributed identifiers and makes use of two
new synchronization objects for managing these identifiers:
a \emph{claim-id object}, used for claiming an available identifier in a given
(bounded or unbounded) set, and a \emph{read-write register object}, used for storing the
currently used identifier.
The claim, unclaim, and check-claim operations can be implemented individually without the Openflow
bundle feature.

@@

The synchronization mechanisms presented in this paper
are based on Openflow~\cite{of-spec}, the standard SDN protocol today. In this section,
we provide the necessary background.

In a nutshell,
Openflow
is based on a match-action concept: Openflow switches store
rules (installed by the controller) consisting of a match and an
action part. A packet matched by a certain rule will be subject
to the associated action (to be immediately applied now or later).
For example, an action can define a port to which the
matched packet should be forwarded, or add or change a tag
(a certain part in the packet header).
Each Openflow switch stores one or multiple flow tables,
each of which contains a set of rules (flow entries). Flow
tables form a pipeline, and flow entries are ordered according
to priorities: A packet arriving at a switch is first checked by
the rule of the highest priority in table 0: the fields of the
data packet are compared with the match fields of that rule,
and if they fit, some instructions (the actions) are executed;
subsequently, lower priority rules are checked. Depending on
the outcome of the table 0 processing, the packet may be sent
to additional flow tables in the pipeline; during the traversal of
the flow tables, the packet can be equipped with meta-tags to
transfer temporary information. Concretely, instructions can be
used to define additional tables to be visited (goto instruction),
to modify the set of to-be-applied actions (either by appending,
deleting, or modifying actions), or immediately apply some
actions to the packet. A meta-data field can be used to exchange
information between tables; it is part of the header that can be
inserted or removed from a packet via push and pop actions.
In general, a packet can be matched against any of its header
fields, and fields can be wildcarded and sometimes bitmasked.
(For instance, the meta-data field is maskable.) If no rule
matches, the packet is dropped.

@@

\textbf{The FlowMod Command.}
\emph{FlowMod} is the main messages used in Openflow 
to update the switch configuration. 
In a nutshell, \emph{FlowMod} specifies the type of flow table modification,
a match and an action. 
The standard behavior is that the new rule simply replaces an existing one
with exactly the same match, and is installed
in addition to an existing one if the match only
partially overlaps. The latter can result
in inconsistencies, especially if the old and the new rule
have the same priority. \stefan{liron, is this correct?}
\emph{FlowMod} can be used together with a the CHECK\_OVERLAP flag
of \emph{FlowMod}, in order to detect conflicts: if the flow spaces of an old
and the new rule overlap, 
that is, if there is an entry with overlapping match part but different action, 
 and if the two rules have the same priority, then 
\emph{FlowMod} fails.

@@

\textbf{Bundling.}
The Openflow bundle can be seen as a ``mini-transaction'':
a bundles aggregates multiple control messages (add, remove, and modify rule)
into one, so that either all of the control messages are
processed, or none of them is. In particular, if, for some reason, the
configuration request contained in one of the bundled messages cannot
be processed, all the other requests  are rejected and an error
message is sent to the controller that issued them.
This bundled processing can be, when properly configured, done in a
\emph{packet-atomic} way, so that each packet is processed according
to the switch configuration either \emph{before} or \emph{after} the
request was processed.
If used together with the CHECK\_OVERLAP flag, the bundle command also returns a reason 
if the modification failed: the failed command in the bundle:
a CHECK\_OVERLAP flag can be defined for each command in the bundle.

@@

@@

\begin{figure*}[t]
\centering
\includegraphics[width=0.8\columnwidth]{loadbal.pdf}~~
\includegraphics[width=0.8\columnwidth]{pic.pdf}\\
\caption{\emph{Left:} Load-balancing gone wrong. \emph{Right:} Illustration of transactional abstraction which  allows
controllers to modify switch configurations in complex ways,
and
depending on other flow spaces.}\label{fig:illu}
\end{figure*}

@@

what is the poewr of bundling flowmods:
- no cookies: delete entries now by specifying exact match part

- alterantive impelmentation without: group is bounded resources, can have own table etc. dont have many controlelrs anyway so not much is needed; also maybe other policies need groups

Before presenting our transactional interface and implementation,
let us first discuss the limitations of the Openflow features discussed above.
At first sight, the bundle feature seems to provide everything
we need: it defines a sequence of read and write operations
 on the switch state which are executed atomically.
 However, such bundle transactions come with many
 limitations. A bundle transaction does not allow
 to change the switch state in any non-trivial way,
 as the read and write actions in the bundle do
 not support any conditional modifications or computations which
 depend on the read results. By leveraging
 the check overlap feature, it is at least possible
 to abort transactions depending on certain conditions
 in the overlapping
 flow spaces. However, in many cases, for example
 a load-balancing application,
 it can be interesting to define more general
 dependencies, which depend on, from a logical perspective,
 seemingly
 completely independent flow spaces,
 which however share the underlying physical network.
In other words, the bundle feature only allows us to
define ``one-shot transactions'',
transactions which do not support any further controller interactions.
Given
the limited computational options the Openflow interface of a switch offers,
it is difficult to implement more complex behavior.

@@

%\section{The Container Object}\label{sec:vision}
\section{-NEW- Transactional Synchronization}\label{sec:main}

This section describes how to introduce more powerful transactions.
Our implementation is based on two new synchronization objects,
allowing to coordinate controllers using identifiers.

At a high level, we export a
transactional interface which allows each controller $i\in[N]$ ($N\geq
2$) to perform multiple
operations on the configuration of an OpenFlow switch as an atomic
\emph{transaction}, inspired by the transactional-memory paradigm~\cite{stm-st95,tm-book}. A transaction may \emph{commit}, in
which case it appears as executed sequentially, or \emph{abort}, in
which none of its operations takes effect.

@@

We assume transactional access to  two kinds of objects: a read-write \emph{memory} abstraction
object, which we later use for storing the current used policy identifier, and a
\emph{resource-claimer} object, provides claiming a resource id (e.g. policy \emph{identifier}) in a
given (bounded or unbounded) set.
In theory one can use the memory abstraction to implement the resource claimer, however we found it far more simple and efficient to consider the resource claimer as base object.
%

@@

\begin{itemize}
\item by performing $\textit{Write}(addr,k)$,
  $addr,k\in\Nat$, the value at memory offset $addr$ is set to $k$;

@@

Note that reading the memory can be done by just reading the switch configuration however such action doesn't affect the execution of the transaction (as $\textit{\compare}$ does).

An immediate extension of the memory interface is the \emph{compare-and-set} (CAS) operation that takes two integer parameters
$\textit{old}$ and $\textit{new}$, reads the register and, if the
value of it is $\textit{old}$, replaces it with $\textit{new}$ and
returns $\textit{true}$; otherwise, $\textit{false}$ is returned and
we say that the CAS operation \emph{fails} (which causes an abort if
the transaction invoking it).

@@

\subsection{Interface}\label{sec:t-if}

@@

We assume that transactions may comprise operations on the flow table
of an SDN switch. Our transactional abstraction exports the
$\textbf{atomic}\{\;\ldots\;\}$ construct
containing a sequence of transactional
operations and may \emph{commit} and return responses to all its operations
or \emph{abort} (so that none of its operation takes effect).
We guarantee that in every execution, all committed
transactions constitute a \emph{legal} sequential
order~\cite{Pap79-serial}, i.e., every transaction appears as executed
sequentially. On the progress side, we ensure that a
transaction $T$ is only allowed to abort if either another
transaction updated a flow table entry read by $T$
or if one of $T$'s operations (\textit{check} or CAS) fails.
We employ the \emph{deferred-update} semantics:
no transaction can affect other transactions before it commits.
In particular, if a controller crashes before committing its
transactions, no future transaction can be affected.
%Note that
%, assuming such a transactional memory,
%a $\emph{contains}(x)$ operation can be implemented as a transaction
%composed of $\textit{cond-add}(x)$ followed by $\textit{cond-delete}(x)$.
%A \emph{set} abstraction is a special instance of a multi-set that is
%only modified with \textit{cond-add} and \textit{cond-delete}
%operations (this way the set only contains at most one copy of a given
%item).

\subsection{Implementation in Openflow}\label{sec:t-impl}

@@

\begin{algorithm}[h]
    \caption{$\textit{try-commit}()$}
    \label{alg:commit}
    \begin{algorithmic}[1]
    \Require  current bundle id $bid$.
		    \State $cmd \gets struct ofp_bundle_ctrl_msg$
    		\State  $cmd.bundle\_id \gets bid$
    		\State  $cmd.type \gets $ \textsf{OFPBCT\_COMMIT\_REQUEST}
    		\State  $cmd.flags \gets $ \textsf{OFPBF\_ATOMIC}
    		\State send $cmd$
		\Return $cmd$
    \end{algorithmic}
\end{algorithm}

@@

For brevity, we omit explicitly putting the $\textbf{atomic}\{...\}$ construct
around each of the transactional operation used in the algorithm.
Indeed, since \textit{claim} and \textit{unclaim} operations cannot
cause the corresponding transaction, we can assume that all
``singleton'' (consisting of one operation only)  transactions
containing such operations always commit.

@@

%\paragraph{One-Shot Update} FIXME: desribe how to implement one-shot update

\paragraph{STN}

\textbf{TODO}

\section{In-Band Composition}\label{sec:compo}

>>>>

\algblockdefx[Bundle]{startTxn}{endTxn} %
[0]{\textbf{bundle}\{} %
[1][res]{ \} $\rightarrow #1$}

@@

\begin{abstract}
Control planes of modern Software-Defined Networks (SDNs)
are \emph{distributed systems}: to ensure availability and fault-tolerance,
to improve load-balancing, and to reduce overheads,
control modules are physically distributed.
%a redundant control plane
%can ensure a high availability, and a spatially distributed
%control plane allows to handle time-critical dataplane
%events close to their origins.
%[[PK this is a contraversial point, not important for us
%Moreover, an SDN may be managed by multiple administrators simultaneously.
%]]
However, a distributed control plane also introduces
new challenges, and in order to guarantee a consistent network operation,
the actions of different controllers may need to be synchronized.
%[[PK sounds contradicting to the fact that such systems already exist
%However, not much is known today about how to implement
%such a distributed control plane.
%]]
%
A well-known concept to solve synchronization and consistency problems
in distributed systems are \emph{transactions}: a transaction defines a sequence
of operations which are performed in an atomic, serializable manner, providing all-or-nothing semantics.
This paper shows that powerful transactions can also be implemented in software-defined networks,
using the standard Openflow protocol.
%Indeed, recent Openflow versions already come with useful but primitive
%features, in particular \emph{bundling} and \emph{rule-overlap checks}.
In particular, we present a
transactional interface which supports arbitrary atomic
transactions
over the switch configuration space: Our transactions
can include complex notions of conflicts (beyond flow overlap checks),
as well as computations and multiple controller interactions. We discuss different use cases
for our interface (consistent policy installation, load-balancing, consensus),
and also describe how to
perform basic forms of policy compositions in-band.
\end{abstract}

@@

While the logically centralized perspective offered by SDN is attractive,
there is a wide consensus that
the control plane should be physically \emph{distributed}.
First, in order to provide a high availability as well as a minimal degree of
fault-tolerance, controllers should be redundant~\cite{onix,stn,onos}: a failure
of one controller can be masked by other controllers. Second, it has also been proposed
to distribute controllers \emph{spatially}, in order to handle latency-sensitive and
communication intensive control plane events close to their origin.~\cite{devoflow,kandoo,jukka,disco}
Third, larger SDNs are likely to be operated by multiple administrators or may even offer
participatory interfaces where different users can install and trigger policy changes
concurrently~\cite{participatory,stn}.

@@

Today, we do not have a good understanding yet of how to realize
such distributed control planes. The problem is essentially a
distributed systems
one: Multiple controllers may simultaneously try to
install conflicting updates and we want to resolve these conflicts
\emph{consistently} (no undesired behavior is observed on the data
plane) and \emph{efficiently} (no undesired delays are imposed on the
control application).
\emph{Synchronizing}
the distributed controllers and manipulating the network state \emph{consistently}~\cite{cpc}.
are non-trivial tasks~\cite{cap-theorem}.

@@

Being able to provide more complex and atomic sequences of transactions over this configuration
space can be appealing to resolve synchronization problems.
This paper investigates the implementation of a transactional interface 
(\`{a} la STN~\cite{stn}) modified concurrently by multiple controllers.

\paragraph{Our Contributions}
This paper shows how to implement a very powerful transactional interface
 for software-defined networks. 
 Concretely, our implementation is based on
 standard Openflow~\cite{of-spec} primitives, which by themselves are very limited.
For example, we find that while Openflow's \emph{bundle} or \emph{rule overlap check}
features can serve as synchronization primitives,
without further mechanisms, their use is very limited:
with a bundle alone it is not possible to react to
and change switch configuration state in any non-trivial way,
and with the rule overlap check it is not possible to react to
update conflicts defined over non-overlapping flow spaces.
Moreover, we show that many powerful synchronous operations
can also be implemented without bundles.

The transactional interface presented in this paper supports very general atomic
transactions over the switch configuration space, similar to classic transactional memory~\cite{stm-st95,tm-book}.
For instance, a transaction allows to modify switch configurations in an interactive
and computational fashion, as well as to define complex notions
of conflicts.
Our implementation is based on distributed identifiers and makes use of two
new synchronization objects for managing these identifiers:
a \emph{claim-id object}, used for claiming an available identifier in a given
(bounded or unbounded) set, and a \emph{read-write register object}, used for storing the
currently used identifier.
The claim, unclaim, and check-claim operations can be implemented individually without the Openflow
bundle feature.

@@

The synchronization mechanisms presented in this paper
are based on Openflow~\cite{of-spec}, the standard SDN protocol today.
% In this section,
%we provide the necessary background.

In a nutshell,
Openflow
follows a match-action concept: Openflow switches store
rules (installed by the controller) consisting of a match and an
action part. A packet matched by a certain rule will be subject
to the associated action (to be immediately applied now or later).
For example, an action can define a port to which the
matched packet should be forwarded, or add or change a tag
(a certain part in the packet header).
Each Openflow switch stores one or multiple flow tables,
each of which contains a set of rules (flow entries). Flow
tables form a pipeline, and flow entries are ordered according
to priorities: A packet arriving at a switch is first checked by
the rule of the highest priority in table 0: the fields of the
data packet are compared with the match fields of that rule,
and if they fit, some instructions (the actions) are executed;
subsequently, lower priority rules are checked. Depending on
the outcome of the table 0 processing, the packet may be sent
to additional flow tables in the pipeline; during the traversal of
the flow tables, the packet can be equipped with meta-tags to
transfer temporary information. Concretely, instructions can be
used to define additional tables to be visited (goto instruction),
to modify the set of to-be-applied actions (either by appending,
deleting, or modifying actions), or immediately apply some
actions to the packet. A meta-data field can be used to exchange
information between tables; it is part of the header that can be
inserted or removed from a packet via push and pop actions.
In general, a packet can be matched against any of its header
fields, and fields can be wildcarded and sometimes bitmasked.
(For instance, the meta-data field is maskable.) If no rule
matches, the packet is dropped.

@@

\subsection{The FlowMod Command}

\emph{FlowMod} is the main messages used in Openflow 
to update the switch configuration. 
In a nutshell, \emph{FlowMod} specifies the type of flow table modification,
a match and an action. 
The standard behavior is that the new rule simply replaces an existing one
with exactly the same match, and is installed
in addition to an existing one if the match only
partially overlaps. The latter can result
in inconsistencies, especially if the old and the new rule
have the same priority. \stefan{liron, is this correct?}
\emph{FlowMod} can be used together with a the CHECK\_OVERLAP flag
of \emph{FlowMod}, in order to detect conflicts: if the flow spaces of an old
and the new rule overlap, 
that is, if there is an entry with overlapping match part but different action, 
 and if the two rules have the same priority, then 
\emph{FlowMod} fails.

@@

\subsection{Bundling}

The Openflow bundle can be seen as a ``mini-transaction'':
a bundles aggregates multiple control messages (add, remove, and modify rule)
into one, so that either all of the control messages are
processed, or none of them is. In particular, if, for some reason, the
configuration request contained in one of the bundled messages cannot
be processed, all the other requests  are rejected and an error
message is sent to the controller that issued them.
This bundled processing can be, when properly configured, done in a
\emph{packet-atomic} way, so that each packet is processed according
to the switch configuration either \emph{before} or \emph{after} the
request was processed.
If used together with the CHECK\_OVERLAP flag, the bundle command also returns a reason 
if the modification failed: the failed command in the bundle:
a CHECK\_OVERLAP flag can be defined for each command in the bundle.

@@

In the remainder of this paper, we will use the notation 
$\textbf{bundle}\{\;\ldots\;\}$ to refer to represent a
sequence of commands in a bundle.
\stefan{liron: i think the following needs more details $:-)$ :} A bundle will always have the following structure:
it first contains a \emph{send message} to start bundle,
followed by a sequence of messages wrapped in bundle and bundle ID, 
and finally end bundle control message for commit.
A bundle always returns a binary result: \emph{error} or \emph{success}.

Algorithm~\ref{alg:commit} describes how to commit
the bundle. 
\begin{algorithm}[h]
    \caption{$\textit{try-commit}()$}
    \label{alg:commit}
    \begin{algorithmic}[1]
    \Require  current bundle id $bid$.
		    \State $cmd \gets struct ofp_bundle_ctrl_msg$
    		\State  $cmd.bundle\_id \gets bid$
    		\State  $cmd.type \gets $ \textsf{OFPBCT\_COMMIT\_REQUEST}
    		\State  $cmd.flags \gets $ \textsf{OFPBF\_ATOMIC}
    		\State send $cmd$
		\Return $cmd$
    \end{algorithmic}
\end{algorithm}

@@

\begin{figure*}[t]
\centering
\includegraphics[width=0.8\columnwidth]{loadbal.pdf}~~
\includegraphics[width=0.8\columnwidth]{pic.pdf}\\
\caption{\emph{Left:} Load-balancing gone wrong. \emph{Right:} Illustration of transactional abstraction which  allows
controllers to modify switch configurations in complex ways,
and
depending on other flow spaces. Colors indicate different actions.}\label{fig:illu}
\end{figure*}

Finally, we remark that our solution is only based on flow tables
and does not make use
of group tables. 
Moreover, we do not make use of cookies: rules
are always identified by exact matches.

@@

\stefan{liron, maybe we can discuss this section together: we need some more ideas}

Before presenting our transactional interface and implementation,
let us first discuss the limitations of the Openflow features discussed above.
At first sight, the bundle feature seems to provide everything
we need: it defines a sequence of read and write operations
 on the switch state which are executed atomically.
 However, such bundle transactions come with several
 limitations, as discussed in the following. Moreover,
 as we will see, our mechanism also does not \emph{need}
 the bundle feature.

A bundle does not allow
 to change the switch state in any non-trivial way,
 as the read and write actions in the bundle do
 not support any conditional modifications or computations which
 depend on the read results. By leveraging
 the check overlap feature, it is at least possible
 to abort transactions depending on certain conditions
 in the overlapping
 flow spaces. However, in many cases, for example
 a load-balancing application,
 it can be interesting to define more general
 dependencies, which depend on, from a logical perspective,
 seemingly
 completely independent flow spaces,
 which however share the underlying physical network.
In other words, the bundle feature only allows us to
define ``one-shot transactions'',
transactions which do not support any further controller interactions.
Given
the limited computational options the Openflow interface of a switch offers,
it is difficult to implement more complex behavior.

@@

%\section{The Container Object}\label{sec:vision}
\section{Transactional Synchronization}\label{sec:main}

This section describes how to introduce more powerful transactions.
Our implementation is based on two new synchronization objects,
allowing to coordinate controllers using identifiers.
At a high level, we export a
transactional interface which allows each controller $i\in[N]$ ($N\geq
2$) to perform multiple
operations on the configuration of an OpenFlow switch as an atomic
\emph{transaction}, inspired by the transactional-memory paradigm~\cite{stm-st95,tm-book}. A transaction may \emph{commit}, in
which case it appears as executed sequentially, or \emph{abort}, in
which none of its operations takes effect.

@@

We implement transactional access to  two kinds of objects: a read-write \emph{memory} abstraction
object, which we later use for storing the current used policy identifier, and a
\emph{resource-claimer} object, provides claiming a resource id (e.g. policy \emph{identifier}) in a
given (bounded or unbounded) set.
In theory one can use the memory abstraction to implement the resource claimer, 
however we found it far more simple and efficient to consider the resource claimer as base object.
%

@@

\begin{itemize}
\item By performing $\textit{Write}(addr,k)$,
  $addr,k\in\Nat$, the value at memory offset $addr$ is set to $k$. For
  example, the offset $addr$ identifies the policy (identifier), 
  but our approach is more general.

@@

Note that reading the memory can be done by just reading the switch configuration however such action doesn't affect the execution of the transaction (as $\textit{\compare}$ does).
As we will see, this abstraction can be implemented
without using the notion of bundles.

An immediate consequence of this memory interface is 
that we can implement a \emph{compare-and-set} (CAS) operation that takes two integer parameters
$\textit{old}$ and $\textit{new}$, reads the register and, if the
value of it is $\textit{old}$, replaces it with $\textit{new}$ and
returns $\textit{true}$; otherwise, $\textit{false}$ is returned and
we say that the CAS operation \emph{fails} (which causes an abort if
the transaction invoking it).

@@

Again, none of these individual operations 
requires bundles.

\begin{comment}
\subsection{Interface}\label{sec:t-if}

@@

We assume that transactions may comprise operations on the flow table
of an SDN switch. Our transactional abstraction exports the
$\textbf{bundle}\{\;\ldots\;\}$ construct
containing a sequence of transactional
operations and may \emph{commit} and return responses to all its operations
or \emph{abort} (so that none of its operation takes effect).
We guarantee that in every execution, all committed
transactions constitute a \emph{legal} sequential
order~\cite{Pap79-serial}, i.e., every transaction appears as executed
sequentially. On the progress side, we ensure that a
transaction $T$ is only allowed to abort if either another
transaction updated a flow table entry read by $T$
or if one of $T$'s operations (\textit{check} or CAS) fails.
We employ the \emph{deferred-update} semantics:
no transaction can affect other transactions before it commits.
In particular, if a controller crashes before committing its
transactions, no future transaction can be affected.
%Note that
%, assuming such a transactional memory,
%a $\emph{contains}(x)$ operation can be implemented as a transaction
%composed of $\textit{cond-add}(x)$ followed by $\textit{cond-delete}(x)$.
%A \emph{set} abstraction is a special instance of a multi-set that is
%only modified with \textit{cond-add} and \textit{cond-delete}
%operations (this way the set only contains at most one copy of a given
%item).

\subsection{Implementation}\label{sec:t-impl}

@@

@@

For brevity, we omit explicitly putting the $\textbf{bundle}\{...\}$ construct
around each of the transactional operation used in the algorithm.
Indeed, since \textit{claim} and \textit{unclaim} operations cannot
cause the corresponding transaction, we can assume that all
``singleton'' (consisting of one operation only)  transactions
containing such operations always commit.

@@

Given the CAS, we can also implement more complex operations,
for example the replicated state machine postulated in STN~\cite{stn}:
in order to choose tags in a consistent manner, CAS needs to be enhanced
with a notion of tags.

%\paragraph{One-Shot Update} FIXME: desribe how to implement one-shot update
%
%\paragraph{STN}
%
%\textbf{TODO}
%

\begin{comment}
\section{In-Band Composition}\label{sec:compo}
